---
title: "Danish Kings"
author: "Emmy-Sophie Herzog og Liv Belert Mortensen"
date: "26-5-25"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

#opt cmmd I, echo then the code will not be in the HTML
```{r library, echo = TRUE}
library(tidyverse)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the separator: -> semicolon  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
#the first two outputs use the read.csv function which is incompatible with out spreadsheet, which is semicolon separated. Therefore we will not use these to work with.
kings1 <- read.csv("data/Kongeraekken2.csv")
head(kings1)

kings2 <- read_csv("data/Kongeraekken2.csv", na = "NULL")
head(kings2)

kings3 <- read.csv2("data/Kongeraekken2.csv") #this output is interacting a lot better with the data set because of the csv2 (for semicolon seperated columns) function so we will continue using this output
head(kings3)

kings4 <- read_csv2("data/Kongeraekken2.csv") #this output also interacts with the data set and is using an underscore _ this us usually better to work with because it looks better, but in this case we deemed it unnecessary when the kings3 is also working fine sufficient.
head(kings4)

```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
every function down below besides the dim() function are tidyverse´ functions.
2. What is the result of running `class()` on the `kings` object created with a tidyverse function.
“spec_tbl_df” “tbl_df”, “tbl”, “data.frame” --> gives a vector of class names found in the object
3. How many columns does the object have when created with these different functions?
with using the ncol() and dim() function, I was able to find out that the number of columns are 19 with 56 rows
4. Show the dataset so that we can see how R interprets each column

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <-kings3 

class(kings) 

ncol(kings)

dim(kings)

glimpse(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r - calculating_duration}
# YOUR CODE


kings_duration <- kings %>% 
  mutate(duration = Regering_slut - Regerings_start)

head(kings_duration, na = "NA") %>% 
  select(Monark, duration)
  

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r - figuring_out_the_average_duration}
# YOUR CODE

duration_vector <- kings_duration[["duration"]]
duration_vector <- as.numeric(kings_duration$duration)

average_duration <- mean(duration_vector, na.rm = TRUE)
#the average reign duration is  19,925 years

```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r - creating_duration_column}
# YOUR CODE -> showing kings who ruled above average duration/more than 19 years
kings_filter <- kings_duration %>% 
  filter(duration > 19) %>% 
  select(Monark, duration)
kings_filter

count(kings_filter)

```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r - kings_filtered}
# YOUR CODE

three_kings_days <- kings_filter %>% 
  arrange(desc(duration)) %>% 
  slice(1:3) %>% 
  mutate(three_kings_days_ruled = duration*365) %>% 
  select(Monark, three_kings_days_ruled) # using select to only show the relevant columns in the HTML

three_kings_days

```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r - visualizing_with_ggplot}
# YOUR CODE

library(ggplot2)
ggplot(kings_duration) +
  geom_point(aes(x = Regerings_start, y = duration)) +
  geom_smooth(aes(x = Regerings_start, y = duration)) +
  labs(
    x = "Start of Reign",
    y = "Length of Reign (years)" 
  )+
  scale_x_continuous(limits = c(958, 2025))




```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!